The impact of earthquake cycle variability on neotectonic and paleoseismic slip rate estimates

### 2018-10-14 manuscript/reviews/response_letter.md

---
title: Response to Reviewers, "The impact of earthquake cycle variability on 
neotectonic and paleoseismic slip rate estimates"
author:
  name: Richard Styron
---

## Review by Michael Oskin

@@

![Figure \label{fig}](../figures/nsv_slip_rate_envelopes.pdf)

The differences between the results of this experiment (fixed per-event
displacement of 1 m) and of the numerical experiment performed in the manuscript
are basically that these results are less smooth, but the total variance at any
point in time (*x-axis*) is less. This is simply due to removing the
stochasticity from one of the two variables in the system. The relative
spread in the data and convergence rates are unchanged. This is to be expected
as the variability in the per-event displacement is the same for all recurrence
distributions, so even though it is a random variable in the simulations, it is
not an experimental variable.

What I find the most interesting about this experiment is that the fluctuations
in the estimated slip rates show very clearly the mean earthquake cycles once
the noise from the per-event displacements has been removed. It is clear that
these are kind of damped or averaged out after ~7 earthquake cycles.
Nonetheless, though this is a cool pattern to see, I don't think it adds
enough insight to be worth including in the manuscript. I have added the figure
and table that show the slip rate variation through time to a new supplemental
document, accompanied by a brief discussion.

*Changes: New experiment added to new supplemental materials, and brief 
discussion in the main manuscript.*

@@

> [O3] I would also like to a see a more quantitative comparison of COV and a 
> convergence on the mean to back up the assertion that COV of the distribution 
> is more important than the distribution itself.

This comment is addressed in the response to [O18] below.

> [O4] The paper would be improved by a more quantitative, empirical basis and 
> discussion of physical processes that may drive such recurrence behavior.

I don't understand how the work could be more quantitative--it's a purely 
numerical study.

As for making it more empirical, there are several ways to do this:

1. *Using empirical distributions for earthquake recurrence and slip.* There 
   isn't a lot of consensus on empirical (non-parametric) earthquake recurrence 
   distributions; most of the community prefers parametric distributions such 
   as the lognormal, Weibull or Brownian Passage Time distributions. As 
   discussed in the paper by Matthews et al. (2002) that is the most prominent 
   introduction to the Brownian Passage Time distribution, given the very small 
   number of samples for earthquake recurrence that we will have for a given 
   section of faulting, it is impossible to discriminate between these 
   distributions, so going with the lognormal (as I have done in this study) is 
   justifiable on empirical grounds as well as practical ones (it is familiar, 
   implemented in many programming environments, and easy to manipulate). With 
   regards to empirical slip rate distributions, I have added an experiment 
   that uses one, which is explained in more detail in response to comment 
   [O6a] below.

2. *Going through the literature and evaluating studies that claim that slip 
   rates have changed (or have not changed) in light of the work presented 
   here.* I considered this, and in fact a major motivation for me to begin the 
   analysis was my skepticism over some recent literature claiming major slip 
   rate changes over relatively short timespans. However, I opted not to do 
   this in the paper, mainly because I didn't want to pick fights. There is a 
   bit of a paradox here: If I claim that the conclusions in a paper claiming 
   secular slip rate changes (or fluctuations) are actually due to aleatoric 
   variability in earthquake recurrence, I will probably anger those authors 
   and decrease the likelihood that they will consider these results in 
   subsequent work. I'd rather write a more toothless paper that doesn't single 
   out any given researchers, and is therefore a bit easier to swallow by all.

3. *Incorporating measurement uncertainty.* Measurement uncertainty is a very 
   large factor that affects the results of any slip rate measurements, and I 
   fully agree with comment [O6b] that it is in most cases underreported, both 
   in offset measurements (as that comment references) and in geochronologic 
   dating of any sort.  I chose to leave it out of this paper because I really 
   wanted to focus on the aleatoric variability, which is generally 
   *underappreciated* as opposed to *underreported*.

Per a discussion of the physics and mechanics behind recurrence behavior: I 
have added a short discussion, but I don't want to really dig into the topic, 
for two reasons:

@@

1. The intended audience for the paper is not only crustal deformation
   researchers, but others in the seismic hazard community as well--this
   includes engineers, geotechnical workers, analysts in the insurance industry,
   etc. In my experience as a member of this community, many others are
   only interested in these sorts of phenomena to the degree that they are
   consequential and actionable; their intellectual interests are often oriented
   towards their fields of expertise (structural engineering, ground motions,
   human and economic exposure, etc.). I want this paper to be a straightforward
   reference for how to evaluate slip rate data in light of aleatory variability
   that is not tied down in jargon or linked to specific geological or
   geophysical models or ideas that may not stand the test of time. Because of
   how variable and poorly-understood earthquake recurrence and fault
   interaction phenomena are, an in-depth discussion without resolution may well
   be off-putting to much of the audience that I would like to read this paper.

2. I don't think that we have a great understanding of the real mechanisms, 
   yet. There are a variety of mechanisms under consideration (e.g., co- and 
   post-seismic elastic and viscoelastic Coulomb stress changes, stress 
   transients, dynamic triggering, pore fluid pressure fluctuations, 
   fluctuations in the frictional failure threshold on a fault) in addition to 
   actual secular changes in tectonic loading rates. The time-dependent 
   mechanisms (particularly post-seismic processes) often show different 
   behavior with regard to whether they are 'spun-up' and at a dynamic 
   equilibrium, or not. And all of these mechanisms are necessarily linked to 
   uncertainty as to how (and where) faults are loaded to begin with--whether 
   the loading is in the elastic crust, in the viscoelastic/viscous mid or 
   lower crust (in a continuum style), or on a discrete creeping dislocation 
   down-dip of the brittle fault. There is a big range of scientific opinion on 
   all of these questions. As a community we are begging for a big review paper 
   to at least concatenate and organize these ideas and potentially test them 
   or at least sort them into compatible vs. mutually exclusive sets for future 
   testing. But we don't have that right now, so the topic is kind of a big 
   mudhole. I will dip my toe in but I really want to avoid falling in for the 
   purposes of this paper.

*Changes: New experiment with empirical slip distribution, and discussion of 
physical mechanisms behind aleatory recurrence variability.*

> [O5] There is a literature of ideas to draw upon, such as post-seismic fault 
> reloading (Kenner and Simons, 2005), earthquake super cycles (Sieh et al., 
> 2008; Weldon et al., 2004), isolated versus fault-network behavior (Berryman 
> et al., 2012). Some of these ideas are discussed briefly but need more 
> explanation.

I have added a short discussion (two paragraphs) on the topic, but as noted in 
my response to [O4] I don't think that a more full discussion is warranted. I 
want this paper to be a simple, easy-to-digest paper and I think that a long 
and necessarily unsatisfying (as we don't have answers yet) discussion on the 
mechanisms behind recurrence variability will be an obstacle, and many readers 
will just put the paper down.

*Changes: discussion added.*

@@

Yes, and some of these cases are cited in the introduction. However, the 
open-interval problem simply deals with the uncertainty in a single recurrence 
interval (the present one), and not the variability that is present throughout 
all of the closed earthquake intervals that have contributed to the measured 
offset; this larger issue is the topic of the manuscript.

*Changes: None..*

> [O8] Page 1, line 13. It seems odd to characterize uncertainty due to a 
> random distribution as epistemic. Isn’t this unreported aleatory uncertainty?

Aleatory and epistemic uncertainty are not mutually exclusive categories. Much 
epistemic uncertainty results from aleatoric variability, particularly when the 
underlying distributions that characterize the aleatoric variability are not 
known.

This is one of those instances: The framing of the situation is that one has 
made a single slip-rate 'measurement' (net slip / time) without knowledge of 
where the fault is in its earthquake cycle, what the past earthquake history 
is, and what the distributions of slip and recurrence are for that fault to 
begin with. Thus the condition is one of ignorance, i.e. epistemic uncertainty, 
and this section of the study shows how to approximately quantify this 
uncertainty under different assumptions of the slip and recurrence 
distributions.

*Changes: None.*

@@

> [O9] Page 1, line 20: Why is marker in quotes?

I wanted to declare that it was a technical term and not a word that I 
arbitrarily applied to the situation. But this isn't necessary.

*Changes: Quotes removed.*

@@

Ok.

*Changes: Sentence broken.*

> [O12] Page 2, line 13 and other citations:  persistent use of ‘e.g’ after 
> citing only one or two articles is poor form and makes this reader think that 
> the author has not adequately explored the literature.

The use of 'e.g.' denotes that the given citations are not authoritative or 
canonical in the sense that the cited works are where the concepts given are 
first introduced or best developed, as this isn't true. The cited works are 
generally just modern, high quality studies that exemplify the topic at hand.

@@

I don't really care what readers may think of the depth of my scholarship.

For what it's worth, 'e.g.' should be before the references but was placed 
after by a LaTeX bug that I hadn't diagnosed.

@@

> [O13a] Page 3,  line 11.  This is not the correct definition of an 
> exponential / poisson distribution.  There is no prescribed number of events, 
> only a prescribed time-independent probability.

The time-independent probability is the mean rate of events. The mean rate of 
events is the mean number of events that occur within some time interval.

Obviously with finite sample sets (of time, or of events) there will be some 
variation--otherwise I wouldn't have written the paper.

Nevertheless, the statement actually made is that the spacing between uniform 
random samples in some interval is characterized by an exponential 
distribution, which is true. It is not stated that this is the definition of 
the distribution.

*Changes: None.*

@@

> [O13b] It is also worth noting that this is physically unrealistic at short 
> time intervals because it violates elastic rebound.

Elastic rebound is a hypothesis, not a law, and is phenomenological instead of 
physical in nature. It is unfortunately a step removed from the modern 
understanding of the mechanics of earthquakes, which are based around stress, 
not strain. These map to each other nicely in the case of elastic and Newtonian 
viscoelastic rheologies, but not as nicely with rate-dependent rheologies, 
which are often considered the best characterization of the lower crust and 
upper mantle (e.g., Hetland and Hager, 2006). It's also hard to put strain in a 
framework with friction, for example. 'Physically realistic' modeling has to 
make a lot of assumptions and use heavy duty equipment (finite elements, for 
example) to incorporate strain.

There is also a separate issue with elastic rebound: It's not very easy to tell 
whether all the accumulated shear strain was released in an earthquake or not. 
What kind of measurements would tell us this?

@@

I strongly suspect that we are fundamentally underestimating the frequency of 
very short recurrence intervals on faults. They're close to invisible to 
paleoseismology, which is our main source of data for recurrence interval 
statistics, because very closely-spaced events may not each produce 
differentiable colluvial wedges or other signs of surface deformation. This 
could plausibly result in a strong sample bias in the statistics. Nonetheless, 
we have clear observations of short recurrence intervals in the past few years. 
For example, some parts of the Monte Vettore fault slipped about 20 cm in the 
Amatrice earthquake and then ~2 m in the Norcia earthquake a few months later 
(Gruppo di Lavoro INGV sul Terremoto di Amatrice-Visso. (2016, October 29). 
PRIMO RAPPORTO DI SINTESI SUL TERREMOTO DI VISSO ML 5.9 DEL 26 OTTOBRE 2016 
(ITALIA CENTRALE). Zenodo. http://doi.org/10.5281/zenodo.163818).

From a fault mechanics perspective, some researchers (for example Mark Zoback 
and his students, primarily Townend, as well as myself) believe that the shear 
stress on a fault required to initiate failure is much greater than the stress 
drop during the event, i.e. shear stress does not go to zero. Failure is 
decently described by Mohr-Coulomb models, and at, say, 10 km depth, the 
confining pressure is almost 300 MPa. With a reasonable rock density (2700 
kg/m$^3$), coefficient of static friction (say 0.5), and pore fluid pressure 
(say 0.3 times lithostatic pressure), the shear stress at failure is 94.5 MPa. 
Stress drops are generally on the order of 0.1-10 MPa (see Peter Shearer's work 
on Brune-type stress drop estimates, for example). So if less than 10% of shear 
stress is relieved in an earthquake, what are the implications for the elastic 
rebound hypothesis?

My take on this is that elastic rebound is a great way to describe the 
phenomenology of earthquakes to a non-geologist. Scientifically, it was an idea 
of absolute genius in 1910, but it isn't a thorough or mechanically sound 
framework for earthquake science a century later. Stress-based frameworks are 
much better suited to both conceptual and quantitative treatments.

*Changes: None.*

@@

> [O14] Page 3, line 24. It would be useful to briefly discuss how shape and 
> scale affect distributions generally. Shape governs the how tailed and is 
> dimensionless; scale determines the spread of the distribution and is 
> dimensioned (in years for this case).

Good idea.  I added brief definitions of the scale, shape and location 
parameters to an earlier paragraph where the terms are first written.

*Changes: definitions added.*

@@

> [O15] Page 4, line 6.  Pareto distribution is another, simpler distribution 
> needing only shape and scale to describe COV > 1

The Pareto distribution isn't used to describe earthquake recurrence, as far as 
I know. I believe that its only use in seismology is the tapered Pareto 
distribution for magnitude-frequency distributions by Yan Kagan (and perhaps 
others); this is not a similar-enough use to include here.

*Changes: None.*

@@

If I say a fault has a mean slip of 1, and a mean recurrence interval of 1, and 
a mean slip rate of 1, it's hard to picture such a fault. Are those reasonable 
values for these parameters, individually? Can't say. Furthermore, the slip 
rate isn't even 1, it's 1/1, because it's a rate and the non-dimensional units 
are still not arithmetically compatible units--multiplication and division are 
possible (kind of, but not reducible) but addition and subtraction are not at 
all defined. (Please forgive me for not knowing enough measure theory to 
rigorously state this...).

Non-dimensionalization in many geological problems reduces the clarity of the 
analysis or solution because it strips it of context. At the same time it can 
facilitate the manipulation of the equations within the study, or during 
programming, etc. Physicists like it because it makes their work easier, I 
would say, and they're all pretty used to it.

Compare this to the dimensionalized problem in the paper. Can you imagine a 
fault with a mean per-event slip of 1 m, a recurrence interval of 1000 years, 
and a slip rate of 1 mm/yr?  I can, and I can place it--it's a small but pretty 
active intraplate fault, or a splay in a plate boundary. It's the kind of fault 
one might actually go and trench.

However, all the numbers are 1 (or 1000) so that it is easy to scale to other 
faults by multiplying by non-dimensional scaling factors (i.e., for a Mongolian 
fault with a slip rate of 3 mm/yr and a per-event slip of 10 m, you scale the 
slip by 10 and the slip rate by 3). It's not accidental that I have chosen 
these numbers, and I have explicitly described how to do the scaling.

*Changes: None.*

@@

> [O18] Page 5, line 15. The statement ‘appears to be related to COV’ is 
> disappointing. Given that this paper is entirely simulation, the author 
> should be able to make a quantitative comparison of slip-rate variance to 
> COV.

Sorry to disappoint.

A comparison between COV and the slip-rate variance at any time is actually 
done in the study (it is in fact the core of the study)--I vary the COV of a 
single distribution (the lognormal distribution) while changing nothing else, 
then I keep the lognormal distribution with a COV of 1 and compare that to a 
very different distribution (the exponential distribution) with a COV of 1. The 
results, which are described unambiguously in the study, show that the slip 
rate variance changes with COV but isn't much affected by the shape of the 
distribution (i.e. lognormal vs. exponential).

There is no reason to pursue this farther here. The work that I have done here 
has covered the (small) space of geologically reasonable distributions; there 
aren't huge gaps left unexplored.

There are two main families of distributions that are in consideration for 
earthquake recurrence: Exponential distributions (possibly with modifications 
such as a stretched or hyperexponential distribution) and lognormal-like 
distributions (lognormal, Wiebull, BPT, etc., which are not distinguishable in 
real paleoseismic datasets). I have compared both types of distributions for a 
single COV and the differences are very minor. There are no alternate 
distribution families in any consideration within seismology that are *more* 
different than these two families.

This work is not meant to offer a mathematical proof, which is why I gave the 
soft 'appears to be related to...' statement instead of a more firm 'is 
demonstrated to be caused by...' or 'is proven to be a function of...'. 
Demonstrating some correlation or relation is enough to further the general 
point, which is that more earthquake recurrence variability will lead to more 
short-term slip rate variability (which seems self-evident in any case).

Going farther would either mean doing many more numerical experiments (which do 
not offer real proof and would just clog the paper) or invoking more 
complicated mathematical tools such as stochastic calculus, which I don't 
really know and am unwilling to teach myself for this paper.

*Changes: None.*

@@

> [O19] Page 5, line 29. This problem has been studied (Weldon et al., 2004; 
> Sieh et al., 2008). The Sumatran subduction zone work is particularly 
> relevant and completely overlooked here.

The work by Weldon et al. (2004) is referenced on Page 6, Line 3. Neither of
these papers deal with the topic of autocorrelated recurrence intervals in any
quantitative or otherwise explicit manner, and I don't read anything that I can
interpret as a qualitative discussion either. One can read both papers and not
get a sense of whether a short recurrence interval implies the next recurrence
interval will be short or long, much less any quantification. Both papers deal
with the concept of 'earthquake supercycles' or groups of earthquakes that are
relatively tightly-spaced and separated from other groups by long recurrence
intervals. This may share a conceptual link, but from a technical perspective
this likely has more to do with periodicity than autocorrelation, and these are
mathematically separate such that a periodic sequence may have positive or
negative autocorrelation. Goh and Barabasi (2008, Europhysics Letters) is a
useful discussion on the topic of autocorrelation vs. periodicity in regards to
quantifying clustering behavior (though one may safely ignore the references to
seismicity in that paper).

[Note: I've analyzed Weldon's data (from K. Scharer's refined OxCal earthquake 
dates) and found that earthquake recurrence interval duration at the Wrightwood 
and Pallett Creek sites has a negative autocorrelation, i.e. a short recurrence 
interval is likely to be followed by a long recurrence interval, and vice 
versa. This is quite unlike the autocorrelation in the Puget Lowlands of
Washington State, which is a network of generally similar faults; here 
autocorrelation is positive, so a long recurrence interval will more likely be 
followed by another long recurrence interval. I don't know what this means, or 
whether it's all noise, but it's intriguing to me and I put this bit in the 
paper in hopes of catching the attention of others who may be looking for a 
project.]

@@

> [O20] Page 6, line 1. Zero friction at rupture arrest is very unrealistic, 
> and not a prerequisite for characteristic behavior.

### 2018-10-14 manuscript/eq_cycle_slip_rates.tex
finished revisions

\begin{abstract}
Because of the natural variability (aleatoric uncertainty) in earthquake
recurrence intervals and coseismic displacements on a fault, cumulative
slip on a fault does not increase linearly or perfectly step-wise with
time; instead, some amount of variability in shorter-term slip rates
results. Though this variability could greatly affect the accuracy of
neotectonic (i.e., late Quaternary) and paleoseismic slip rate
estimates, these effects have not been quantified. In this study,
idealized faults with four different, representative earthquake
recurrence distributions are created with equal mean recurrence
intervals (1,000 years) and coseismic slip distributions, and the
variability in slip rate measurements over 500 to 100,000 year
measurement windows is calculated for all faults through Monte Carlo
simulations. The recurrence distributions used are quasi-periodic,
unclustered and clustered lognormal distributions, and an unclustered
exponential distribution. The results demonstrate that the most
important parameter is the coefficient of variation (\(COV\) = standard
deviation / mean) of the recurrence distributions rather than the shape
of the distribution itself. Slip rate variability over short time scales
(\textless{} 10,000 years or 10 mean earthquake cycles) is quite high,
but decreases with time and is close to stable after
\textasciitilde{}40,000 years (40 mean earthquake cycles). This
variability is higher for recurrence distributions with a higher
\(COV\). The natural variability in the slip rate estimates compared to
the true value is then used to estimate the epistemic uncertainty in a
single slip rate measurement (as one would make in a geological study)
in the absence of any measurement uncertainty. This epistemic
uncertainty is very high (a factor of 2 or more) for measurement windows
of a few mean earthquake cycles (as in a paleoseismic slip rate
estimate), but decreases rapidly to a factor of 1-2 with \textgreater{}5
mean earthquake cycles (as in a neotectonic slip rate study). These
uncertainties are independent of, and should be propagated with
uncertainties in fault displacement and geochronologic measurements used
to estimate slip rates. They may then aid in the comparison of slip
rates from different methods or the evaluation of potential slip rate
changes over time.
\end{abstract}

@@

Fault slip rates are generally estimated by dividing measurements of the
offset of geologic marker features by the time over which that offset
accumulated. The uncertainty in the resulting slip estimate is typically
treated as \emph{epistemic}, and quantified through the propagation of
the measurement uncertainties on the offset and time quantities
\citep[e.g.]{bird_uncertainties_2007,zechar_incorporating_2009}.
However, for slip rate estimates on active faults made from offset
measurements near the fault trace (i.e., within a horizontal distance
that is a small fraction of the fault's locking depth), the episodic
nature of surface displacement due to the earthquake cycle will
necessarily affect the results: If the measurements are taken
immediately before an earthquake, the measured offset and resulting slip
rate estimate will be lower than average, while if the measurements are
taken immediately after an earthquake, the offset and rate will be
higher.

The magnitude of the perturbation to the slip rate is, of course, a
function of the number of cumulative earthquakes that have contributed
to the measured offset. For older Quaternary markers that have
experienced tens to hundreds of major earthquakes, the effects will be
minor, and for bedrock geologic markers with kilometers of displacement,
the earthquake cycle is likely not worth accounting for. However, due to
progressive erosion of geologic markers and the challenge of dating many
late Pliocene to early Quaternary units (which are too old for
radiocarbon and many cosmogenic nuclide systems), geologists often have
no choice but to choose late Pleistocene to Holocene markers to date,
and these units may also be more desirable targets if the scientists are
primarily concerned with estimating the contemporary slip rate on a
fault with a slip rate that may vary over Quaternary timescales
\citep[e.g.]{rittase_temporal_2014,zinke_highly_2018}. For slow-moving
faults, the slip either long waiting to be released, or recently
released, may represent a sizeable fraction of the measured fault
offset.

Careful paleoseismologists and neotectonicists will take this into
account in their slip rate calculations if sufficient data are
available, especially in the years after a major earthquake
\citep[e.g.]{rizza_earthquake_2015}, and many others will discuss the
potential effects if the data are not \citep[e.g.]{lifton_latest_2015}.
These workers may only consider the time since the last earthquake,
often making the assumption (stated or not) that the earthquakes are
identical in slip and perfectly periodic.

However, the recurrence intervals between successive earthquakes on any
given fault segment have some natural variability (i.e. \emph{aleatoric
uncertainty}); similarly, displacement at a measured point is not
identical in each earthquake \citep[e.g.]{duross_holocene_2008}.
Therefore, the measured slip rate may deviate from the time-averaged
rate based on the amount of natural variability in the earthquake cycle,
particularly given successive events from the tails of the recurrence
interval or displacement distributions.

The physical mechanisms responsible for the aleatoric variability in earthquake
recurrence intervals and displacements are still unclear, and the subject of
active investigation. Most earthquakes serve to release differential stresses
caused by relative motions of tectonic plates or smaller crustal blocks;
relative plate velocities measured over tens of thousands to millions of years
from geologic reconstructions are similar enough to those measured over a few
years through GPS geodesy that sudden, transient accelerations and decelerations
are unlikely (e.g., **DEMETS AND DIXON 1999**). As a consequence, some plate
boundary faults are among the most regularly-rupturing faults known,
particularly sections that are isolated from nearby faults
\citep[e.g.]{berryman_major_2012} and therefore not affected by stress
perturbations resulting from earthquakes on other faults. These stress changes
may be 'static' coseismic instantaneous stresses in the elastic upper crust resulting from
earthquake displacement, or analogous stress changes in the viscoelastic lower
crust or upper mantle from the time-dependent relaxation of static stress
perturbations, or 'dynamic' transient stress changes that accompany the passage
of seismic waves from nearby or distant earthquakes. Regardless of the nature

Intraplate faults

It is important to keep in mind that each earthquake in a sequence places
points of rock on both sides of the fault that were not previously in contact.

@@

In this study, the effects of the natural variability in earthquake
recurrence intervals and per-event displacements on neotectonic slip
rate estimates are investigated through Monte Carlo simulations. The
study is geared towards providing useful heuristic bounds on the
aleatoric and epistemic uncertainty of late Quaternary slip rate
estimates for fault geologists, probabilistic seismic hazard modelers,
and others for whom such uncertainties are important.

@@

The most commonly used is the \emph{exponential} distribution. This is
associated with a Poisson process, and is the distribution that results
from earthquakes being distributed uniformly randomly within some time interval.
Consequently, the probability of an earthquake (or other event) occurring at any
time does not change with time since the previous event (in other words, the
hazard function is time-invariant); this leads to the characterization of the
exponential recurrence distribution as `random', `memoryless' or
`time-independent'. The exponential distribution is also the simplest to
describe statistically, as it requires only one parameter (the mean rate
parameter), which is a statistical scale parameter. The standard deviation of a
large number of samples generated from an exponential distribution is equal to
the mean.

@@

The behavior of these distributions and of empirical datasets may be
characterized by the regularity of the spacing between events (i.e.~the
recurrence intervals): these may be periodic, unclustered (i.e.,
`random'), or clustered. Assignment into these categories is typically
done with a parameter known as the coefficient of variation, or
\(COV = \sigma / \mu\), where \(\sigma\) is the standard deviation of
the recurrence intervals, and \(\mu\) is the mean recurrence interval.

Periodic earthquakes are those that occur more regularly than random,
and have a \(COV < 1\) (i.e. \(\sigma < \mu\)). These may be generated
by any of the time-dependent distributions described above with suitable
scale and shape parameters. (Note that in this paper, the use of the
term `periodic' does not mean \emph{perfectly} repeating as it might in
the physics or mathematics literature.)

Unclustered earthquakes occur as regularly as random, and have a
\(COV=1\). These may be generated by the exponential distribution (which
can generate no other), or by any of the time-dependent distributions as
well, given the appropriate parameters. Note that sample sets generated
from these different distributions will not be identical, however:
Sequences with an exponential recurrence distribution will have many
more pairs of events that are much more closely spaced together than the
mean, and more pairs of events that are much more widely spaced than the
mean, compared to a sequence generated from the lognormal distribution.
Nonetheless, these will cancel out in the aggregate statistics, so that
the standard deviations will be equal. A comparison of these may be seen
in Figure \ref{eq_rec_dists}.

Clustered earthquake sequences have sets of very tightly spaced
earthquakes that are widely separated, and have a \(COV>1\). These may
be generated from a hyperexponential distribution, which is the sum of
multiple exponentials with different means, or from the time-dependent
distributions given above, given the right parameters.

>>>>

\begin{abstract}
Because of the natural (aleatoric) variability in earthquake
recurrence intervals and coseismic displacements on a fault, cumulative
slip on a fault does not increase linearly or perfectly step-wise with
time; instead, some amount of variability in shorter-term slip rates
results. Though this variability could greatly affect the accuracy of
neotectonic (i.e., late Quaternary) and paleoseismic slip rate
estimates, these effects have not been quantified. In this study,
idealized faults with four different, representative earthquake
recurrence distributions are created with equal mean recurrence
intervals (1,000 years) and coseismic slip distributions, and the
variability in slip rate measurements over 500 to 100,000 year
measurement windows is calculated for all faults through Monte Carlo
simulations. The recurrence distributions used are quasi-periodic,
unclustered and clustered lognormal distributions, and an unclustered
exponential distribution. The results demonstrate that the most
important parameter is the coefficient of variation (\(CV\) = standard
deviation / mean) of the recurrence distributions rather than the shape
of the distribution itself. Slip rate variability over short time scales
(\textless{} 5,000 years or 5 mean earthquake cycles) is quite high, varying by 
    a factor of 3 or more from the mean,
but decreases with time and is close to stable after
\textasciitilde{}40,000 years (40 mean earthquake cycles). This
variability is higher for recurrence distributions with a higher
\(CV\). The natural variability in the slip rate estimates compared to
the true value is then used to estimate the epistemic uncertainty in a
single slip rate measurement (as one would make in a geological study)
in the absence of any measurement uncertainty. This epistemic
uncertainty is very high (a factor of 2 or more) for measurement windows
of a few mean earthquake cycles (as in a paleoseismic slip rate
estimate), but decreases rapidly to a factor of 1—2 with \textgreater{}5
mean earthquake cycles (as in a neotectonic slip rate study). These
uncertainties are independent of, and should be propagated with
uncertainties in fault displacement and geochronologic measurements used
to estimate slip rates. They may then aid in the comparison of slip
rates from different methods or the evaluation of potential slip rate
changes over time.
\end{abstract}

@@

Fault slip rates are generally estimated by dividing measurements of the
offset of geologic marker features by the time over which that offset
accumulated. The uncertainty in the resulting slip estimate is typically
treated as \emph{epistemic}, and quantified through the propagation of
the measurement uncertainties on the offset and time quantities
\citep[e.g.,][]{bird_uncertainties_2007,zechar_incorporating_2009}.
However, for slip rate estimates on active faults made from offset
measurements near the fault trace (i.e., within a horizontal distance
that is a small fraction of the fault's locking depth, as the width of the zone 
affected by earthquake-cycle strains is a function of locking depth 
\citep[e.g.,][]{savage_geodetic_1973, hetland_interseismic_2006}), the episodic
nature of surface displacement due to the earthquake cycle will
necessarily affect the results: If the measurements are taken
immediately before an earthquake, the measured offset and resulting slip
rate estimate will be lower than average, while if the measurements are
taken immediately after an earthquake, the offset and rate will be
higher.

The magnitude of the perturbation to the slip rate is, of course, a
function of the number of cumulative earthquakes that have contributed
to the measured offset (plus any aseismic strain such as afterslip or creep).
For older Quaternary markers that have experienced tens to hundreds of major
earthquakes, the effects will be minor, and for bedrock geologic markers with
kilometers of displacement, the earthquake cycle is likely not worth accounting
for. However, due to progressive erosion of geologic markers and the challenge
of dating many late Pliocene to early Quaternary units (which are too old for
radiocarbon and many cosmogenic nuclide systems), geologists often have no
choice but to choose late Pleistocene to Holocene markers to date. These
units may also be more desirable targets if the scientists are primarily
concerned with estimating the contemporary slip rate on a fault with a slip
rate that may vary over Quaternary timescales
\citep[e.g.,][]{rittase_temporal_2014,zinke_highly_2018}. For slow-moving 
faults, the slip either long waiting to be released, or recently released, may 
represent a sizeable fraction of the measured fault offset.

Careful paleoseismologists and neotectonicists will take this into
account in their slip rate calculations if sufficient data are
available, especially in the years after a major earthquake
\citep[e.g.,][]{rizza_earthquake_2015}, and many others will discuss the
potential effects if the data are not \citep[e.g.,][]{lifton_latest_2015}.
These workers may only consider the time since the last earthquake,
often making the assumption (stated or not) that the earthquakes are
identical in slip and perfectly periodic.

However, the recurrence intervals between successive earthquakes on any
given fault segment have some natural (i.e. \emph{aleatoric})
variability; similarly, displacement at a measured point is not
identical in each earthquake \citep[e.g.,][]{duross_holocene_2008}.
Therefore, the measured slip rate may deviate from the time-averaged
rate based on the amount of natural variability in the earthquake cycle,
particularly given successive events from the tails of the recurrence
interval or displacement distributions.

The physical mechanisms responsible for the aleatoric variability in earthquake 
recurrence intervals and displacements are still unclear, and the subject of 
active investigation. Most earthquakes serve to release differential stresses 
caused by relative motions of tectonic plates or smaller crustal blocks; 
relative plate velocities measured over tens of thousands to millions of years 
from geologic reconstructions are similar enough to those measured over a few 
years through GPS geodesy that sudden, transient accelerations and 
decelerations are unlikely \citep[e.g.,][]{demets_new_1999}. As a consequence, 
plate boundary faults may have near-constant rates of loading from tectonic 
stress. Many plate boundary faults are are among the most regularly-rupturing 
faults known, particularly sections that are isolated from nearby faults 
\citep[e.g.,][]{berryman_major_2012} and therefore not affected by stress 
perturbations resulting from earthquakes on other faults.

These stress perturbations may be `static' coseismic instantaneous stresses in 
the elastic upper crust resulting from earthquake displacement 
\citep{king_static_1994}, or analogous post-seismic stress changes in the 
viscoelastic lower crust or upper mantle from the time-dependent relaxation of 
static stress perturbations \citep[e.g.,][]{chery_postseismic_2001}, or 
`dynamic' transient stress changes that accompany the passage of seismic waves 
from nearby or distant earthquakes \citep{gomberg_dynamic_2005}.  Additionally, 
changes in pore fluid pressure in a fault zone may increase or decrease the 
required shear stress to initiate an earthquake 
\citep{steacy_introduction_2005}. In contrast to isolated plate boundary 
faults, intraplate faults or those on distributed plate boundaries may have 
lower stress accumulation rates and the stress perturbations from activity on 
other faults may be enough to significantly affect the timing of earthquakes on 
a given fault \citep{gomberg_time-dependent_2005}.

@@

In this study, the effects of the natural variability in earthquake
recurrence intervals and per-event displacements on neotectonic slip
rate estimates are investigated through Monte Carlo simulations. The
study is geared towards providing useful heuristic bounds on the
aleatoric variability and epistemic uncertainty of late Quaternary slip rate
estimates for fault geologists, probabilistic seismic hazard modelers,
and others for whom such uncertainties are important.

@@

The most commonly used is the \emph{exponential} distribution. This is
associated with a Poisson process, and is the distribution that results
from earthquakes being distributed uniformly randomly within some time interval.
Consequently, the probability of an earthquake (or other event) occurring at any
time does not change with time since the previous event (in other words, the
hazard function is time-invariant); this leads to the characterization of the
exponential recurrence distribution as `random', `memoryless' or
`time-independent'. The exponential distribution is also the simplest to
describe statistically, as it requires only one parameter (the mean rate
parameter), which is the inverse of the statistical scale parameter (the scale 
parameter of a distribution determines the dispersion of the distribution, 
while the shape parameter determines the shape or form of the distribution; a 
parameter that shifts the distribution along the $x$-axis is called a location 
parameter).  The standard deviation of a large number of samples generated from 
an exponential distribution is equal to the mean.

@@

The behavior of these distributions and of empirical datasets may be
characterized by the regularity of the spacing between events (i.e.~the
recurrence intervals): these may be periodic, unclustered (i.e.,
`random'), or clustered. Assignment into these categories is typically
done with a parameter known as the coefficient of variation, or
\(CV = \sigma / \mu\), where \(\sigma\) is the standard deviation of
the recurrence intervals, and \(\mu\) is the mean recurrence interval.

Periodic earthquakes are those that occur more regularly than random,
and have a \(CV < 1\) (i.e. \(\sigma < \mu\)). These may be generated
by any of the time-dependent distributions described above with suitable
scale and shape parameters. (Note that in this paper, the use of the
term `periodic' does not mean \emph{perfectly} repeating as it might in
the physics or mathematics literature; the behavior referred to is most 
accurately termed `quasi-periodic', but that term will not be used in the 
interests of conciseness.)

Unclustered earthquakes occur as regularly as random, and have a
\(CV=1\). These may be generated by the exponential distribution (which
can generate no other), or by any of the time-dependent distributions as
well, given the appropriate parameters. Note that sample sets generated
from these different distributions will not be identical, however:
Sequences with an exponential recurrence distribution will have many
more pairs of events that are much more closely spaced together than the
mean, and more pairs of events that are much more widely spaced than the
mean, compared to a sequence generated from the lognormal distribution.
Nonetheless, these will cancel out in the aggregate statistics, so that
the standard deviations will be equal. A comparison of these may be seen
in Figure \ref{eq_rec_dists}.

Clustered earthquake sequences have sets of very tightly spaced
earthquakes that are widely separated, and have a \(CV>1\). These may
be generated from a hyperexponential distribution, which is the sum of
multiple exponentials with different means, or from the time-dependent
distributions given above, given the right parameters.
